---
title: "Databases With R"
author: "Andrew Bates"
institute: "San Diego R Users Group"
date: "2020-07-30"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
```




```{r pokeapi, echo = FALSE}
knitr::include_graphics("https://raw.githubusercontent.com/PokeAPI/media/master/logo/pokeapi.svg")
```

<br>
<br>

PokeAPI: [https://pokeapi.co/](https://pokeapi.co/)


---

# Connect to a (local) database

```{r connect-to-db}
con <- DBI::dbConnect(
  RSQLite::SQLite(),
  dbname = "data/pokedex.sqlite"
  )
```


---

# Query a table with SQL

```{r query-with-dbi}
DBI::dbGetQuery(
  con,
  "SELECT *
   FROM pokemon
   LIMIT 10"
)

```


---

# Connect to a table with dplyr

```{r pokemon-tbl}
pokemon <- tbl(con, "pokemon")
pokemon
```


---

# Query a table with dplyr

```{r pokemon-query-with-dplyr, warning = FALSE}
pokemon %>% 
  summarise(
    mean_height = mean(height),
    max_weight = max(weight),
    sd_xp = sd(base_xp)
  )
```


---

# Query multiple tables with dplyr

```{r types-tbl}
types <- tbl(con, "types")
types
```


---

# Query multiple tables with dplyr

```{r join}
pokemon_types <- pokemon %>% 
  left_join(types, by = "pokemon_id")

pokemon_types %>% 
  select(name, type)
```


---

class: inverse, center, middle

How does dplyr do this?

---

# How does dplyr do this?


> dbplyr is the database backend for dplyr. It allows you to use remote database tables as if they are in-memory data frames by automatically converting dplyr code into SQL.

* dplyr is the front end
* d*b*plyr is the back end

---

# How does dplyr do this?

```{r dplyr-to-sql}
pokemon %>% 
  show_query()
```


---

# How does dplyr do this?

```{r dplyr-to-sql-join}
pokemon %>% 
  left_join(types, by = "pokemon_id") %>% 
  show_query()
```



---

class: inverse, center, middle

DBI

---

# DBI

> A database interface (DBI) definition for communication between R and RDBMSs 

* defines a databases interface (front end)
* database specific packages define the back end (e.g. RSQLite)

---

class: center, middle

```{r dbi-interfaces, echo = FALSE}
knitr::include_graphics("https://mermaid.ink/svg/eyJjb2RlIjoiZ3JhcGggVERcbiAgQShEQkkpIC0tPiBCKFJTUUxpdGUpXG4gIEEgLS0-IEMob2RiYylcbiAgQSAtLT4gRChSUG9zdGdyZXMpXG4gIEEgLS0-IEUoYmlncXVlcnkpXG4gIEEgLS0-IEYoLi4uKVxuXHRcdCIsIm1lcm1haWQiOnsidGhlbWUiOiJkZWZhdWx0In19")

```


---

class: inverse, center, middle

Tips

---

# dbplyr: non-default schema

```{r dbplyr-in-schema, eval = FALSE}
my_table <- tbl(
  con,
  dbplyr::in_schema("schema_name", "table_name")
)
```


---

# dbplyr: non-R functions

prefix functions - leave as is

```{r dplyr-leave-as-is}
pokemon %>% 
  transmute(name2 = SUBSTR(name, 1, 2))
```

---

# dbplyr: non-R functions

infix functions - use '%'

```{r dplyr-infix}
pokemon %>% 
  filter(name %like% 'ab%')
```

---

# dbplyr: non-R functions

special syntax - use `sql()`


```{r dbplyr-special-syntax}
pokemon %>% 
  transmute(height_int = sql("CAST (height AS INT)"))
```

---

# Database package

* Create helper functions to:

    * connect to different servers
    * connect to commonly used tables

* Much easier than remembering 'database xyz', 'schema blah', 'table table_name_1234'

<br>

```{r db-package-example, eval = FALSE}
library(mycompanydbs)
con <- connect("prod")
customers <- remote_customers(con, year = 2020)
```


---

class: inverse, center, middle

Resources


---

# Resources

* [db.rstudio.com](https://db.rstudio.com/)

* [dbplyr.tidyverse.org](https://dbplyr.tidyverse.org/index.html)

* [dbi.r-dbi.org](https://dbi.r-dbi.org/)


---

# Do more with databases

* Fit models in database with [modeldb](https://modeldb.tidymodels.org/)

* Use a database to do plot calculations with [dbplot](https://edgararuiz.github.io/dbplot/)


---
class: inverse, center, middle

# Thanks!
